Towards Generic Parallel Programming in Computer Science Education with Kokkos

### 2020-10-13 src/2.GenericProgramming.tex
Cam-ready version.

\begin{figure}[h]
\begin{minted}[fontsize=\small]{text}
Constrained Semantics: memory allocations, 
memory movement, execution parameters
Parallel Paradigms: patterns, tasks
Expression: language 
Control Paradigm: prescriptive, descriptive
\end{minted}
\caption{Semantic capture: a set of constrained semantics, parallel paradigms, expression and concern define key properties of a parallel programming model.}
\label{figSemCapture}
\end{figure}

\emph{Constrained Semantics} provided by the programmer to the parallel programming model can be grouped into information to express intent or correctness and properties that control critical performance aspects. They address the question of \emph{what}, \emph{where} and \emph{how}. In the context of parallel programming, these correspond to defining which code portion to parallelize, where to run the code, where to access the data, and how to run that parallel code. Synchronization primitives may be considered part of the \emph{what}, and information on the execution properties such as data placement or memory access type may be considered as part of the \emph{how}. Application logic is expressed following a \emph{paradigm} implemented in a \emph{programming language}.

@@

Both examples provide the semantic information on what to parallelize, however, they do not expose enough abstraction primitives that would allow them to map parallel execution on modern architectures efficiently. To define what further information is needed to express concurrency on modern computer architectures, it is important to define an~\emph{abstract machine representation} first. In the following chapters we use the words \emph{representation} and \emph{model} interchangeably. In this work, the abstract machine model does not describe a computer in terms of automata theory but in terms of physical or functional components.

\begin{figure}
\begin{minted}[fontsize=\small]{c++}
#pragma model parallel loop
for ( size_t i = 0; i < N; ++i) {
 /* loop body */
}
\end{minted}
\caption{Annotations-based programming models extend the semantic of the base language through pragma annotations.}
\label{figOMPLike}
\end{figure}

\begin{figure}
\begin{minted}[fontsize=\small]{c++}
model::parallel_for (N, [=] ( const size_t i) {
  /* loop body */
});
\end{minted}
\caption{Embedded parallel programming uses the base language and relies on programming interfaces to capture information.}
\label{figKokkosLike}
\end{figure}

>>>>

\begin{figure}[t!]
\begin{minted}[fontsize=\small]{text}
Constrained Semantics: memory allocations, 
memory movement, execution parameters
Parallel Paradigms: patterns, tasks
Expression: language 
Control Paradigm: prescriptive, descriptive
\end{minted}
\caption{Semantic capture: a set of constrained semantics, parallel paradigms, expression, and control paradigm define key properties of a parallel programming model.}
\label{figSemCapture}
\end{figure}

\emph{Constrained Semantics} provided by the programmer to the parallel programming model can be grouped into information to express intent or correctness and properties that control critical performance aspects. They address the question of \emph{what}, \emph{where}, and \emph{how}. In the context of parallel programming, these correspond to defining which code portion to parallelize, where to run the code, where to access the data, and how to run that parallel code. Synchronization primitives may be considered part of the \emph{what}, and information on the execution properties such as data placement or memory access type may be considered as part of the \emph{how}. Application logic is expressed following a \emph{paradigm} implemented in a \emph{programming language}.

@@

Both examples provide the semantic information on what to parallelize, however, they do not expose enough abstraction primitives that would allow them to map parallel execution on modern architectures efficiently. To define what further information is needed to express concurrency on modern computer architectures, it is important to define an~\emph{abstract machine representation} first. In the following chapters, we use the words \emph{representation} and \emph{model} interchangeably. In this work, the abstract machine model does not describe a computer in terms of automata theory but in terms of physical or functional components.

\begin{figure}[t!]
\begin{minted}[fontsize=\small]{c++}
#pragma model parallel loop
for ( size_t i = 0; i < N; ++i) {
 /* loop body */
}
\end{minted}
\caption{Annotations-based programming models extend the semantic of the base language through pragma annotations.}
\label{figOMPLike}
\end{figure}

\begin{figure}[t!]
\begin{minted}[fontsize=\small]{c++}
model::parallel_for (N, [=] ( const size_t i) {
  /* loop body */
});
\end{minted}
\caption{Embedded parallel programming uses the base language and relies on programming interfaces to capture information.}
\label{figKokkosLike}
\end{figure}

### 2020-10-13 src/3.MachineModel.tex
Cam-ready version.

The Kokkos abstract machine representation targets a design of a future shared-memory computing architecture. The design is shown in Figure~\ref{fig:spaces}. It is characterized by multiple latency-optimized execution units and off-die bandwidth-optimized accelerators. Both compute device types can have disjoint memory address spaces with unique performance properties. In such an architecture, execution units might be hierarchically organized with multiple levels of achievable parallelism with different memory access characteristics and coherence properties of caches. In order to ensure performance portability on such a wide range of configurations, an abstraction of compute-resources and available memories is required. For this purpose, we introduce the abstract concept called \emph{space}.

@@

An instance of a memory space provides a concrete method for the application programmer to request data storage allocations. Following the architecture as discussed earlier, the multi-core processor may contain multiple memory spaces that abstract on-package memory, slower DRAM, and non-volatile memories. Accelerators can provide an additional memory space through its local on-package memory. This is also highlighted in Figure~\ref{fig:spaces}. The programmer is free to decide where each data structure is allocated by requesting the corresponding memory space. Programmatically this is achieved by instantiating the according memory space. Kokkos provides the appropriate abstraction of memory allocation and data management. We believe that an abstract representation of execution- and memory devices is a key property towards performance-portable parallel programming.

@@

\begin{figure}

>>>>

The Kokkos abstract machine representation targets a design of a future shared-memory computing architecture. The design is shown in Figure~\ref{fig:spaces}. It is characterized by multiple latency-optimized execution units and off-die bandwidth-optimized accelerators. Both compute device types can have disjoint memory address spaces with unique performance properties. In such an architecture, execution units might be hierarchically organized with multiple levels of achievable parallelism with different memory access characteristics and coherence properties of caches. To ensure performance portability on such a wide range of configurations, an abstraction of compute-resources and available memories is required. For this purpose, we introduce the abstract concept called \emph{space}.

@@

An instance of a memory space provides a concrete method for the application programmer to request data storage allocations. Following the architecture, as discussed earlier, the multi-core processor may contain multiple memory spaces that abstract on-package memory, slower DRAM, and non-volatile memories. Accelerators can provide an additional memory space through its local on-package memory. This is also highlighted in Figure~\ref{fig:spaces}. The programmer is free to decide where each data structure is allocated by requesting the corresponding memory space. Programmatically this is achieved by instantiating the according memory space. Kokkos provides the appropriate abstraction of memory allocation and data management. We believe that an abstract representation of execution- and memory devices is a key property towards performance-portable parallel programming.

@@

\begin{figure}[t!]

### 2020-10-13 src/4.ProgammingModel.tex
EduHPC @ SC2020 Camera Ready
Cam-ready version.

Kokkos defines a pure interface for C++ and uses C++ as a programming language. This has multiple reasons. It is our experience that developers of scientific and high-performance computing applications express the demand for C++ and a generic parallel programming model to support their codes. Further, following their estimates, porting efforts towards adapting to vendors, programming models, APIs, and software releases represents a burden worth addressing. Lastly, the possibility to maintain pure C++ codes is appealing from the code development and debugging perspective.

@@

To capture semantic information, that is the~\emph{what}, the~\emph{where} and the~\emph{how}, Kokkos introduces six abstractions:~\emph{execution spaces},~\emph{execution patterns},~\emph{execution policies},~\emph{memory spaces},~\emph{memory layout} and~\emph{memory traits}. These abstractions specify semantic information that enables to capture the programmer's intent and allows the runtime to efficiently map the program to any underlying hardware architecture. We list them as follows:
\begin{itemize}
	\item  An execution space is a place where code can be executed. On current hardware architectures, this corresponds to accelerators and CPUs and can include other compute devices in the future. This abstraction supports remote compute devices in distributed memory scenarios as remote execution spaces.
	\item Execution patterns expose the parallel programming paradigm. Supported patterns are \emph{parallel\_for} loop that executes the loop body in any order a specified amount of times, the \emph{parallel\_reduce} which combines a parallel\_for with a reduction operation, \emph{parallel\_scan} which combines a parallel\_for operation with a prefix or postfix scan, and \emph{task} which executes a single function potentially in parallel in respect to other tasks or code sections. 
	\item Execution policies shape the iteration space of a loop pattern. A simple execution policy is a range policy. It specifies that the loop body is executed once for each element in a range.
	\item Memory spaces are the places where data resides. They specify the physical locations of data as well as access characteristics. Different physical locations correspond to different device types such as high bandwidth memories, on-die scratch memories, or non-volatile bulk storage. Different logical memory spaces allow for concepts such as memory in the CUDA programming model, which is accessible from the host and the CUDA accelerator. Memory spaces, similarly to execution spaces, conceptually support remote memory locations in distributed-memory scenarios. Furthermore, they encapsulate functionality such as consistency control and persistence scopes.
	\item Layouts express the mapping from array indices to address offsets. By adopting appropriate layouts for memory structures, an application can optimize data access patterns in a given algorithm. If an implementation provides polymorphic layouts (i.e. a data structure can be instantiated at compile or runtime with different layouts), architecture-dependent optimizations can be performed.
	\item Memory traits specify how a data structure is accessed. Traits express usage scenarios such as atomic access, random access, and streaming loads or stores. This allows the programming model to optimize load and store operations.
\end{itemize}

\begin{figure}[h]
\begin{minted}[fontsize=\small]{text}
Constrained Semantics: Patterns (intent),
Spaces, Layouts, Policies and Traits
Parallel Paradigms: Parallel patterns and 
tasking
Expression: Embedded, C++
Control Paradigm: descriptive
\end{minted}
\caption{Semantic capture defined by the Kokkos programming model.}
\label{fig:SemCaptureKokkos}
\end{figure}

\begin{figure}
\begin{minted}[fontsize=\small]{c++}
template <class ExecPolicy, class FunctorType>
Kokkos::parallel_for(const std::string& name, 
 const ExecPolicy& policy, 
 const FunctorType& functor);
\end{minted}
\caption{The Kokkos~\emph{parallel\_for} class shows how template metaprogramming allows to specialize types. Its descriptive semantic offers the freedom to optimize execution on modern architectures with multiple degrees of concurrency where an execution policy (\emph{ExecPolicy}) shapes the iteration space accordingly.}
\label{fig:parallelFor}
\end{figure}

@@

@@

\begin{figure}
\centerline{\includegraphics[width=0.49\textwidth]{img/Abstractions.png}}
\caption{Overview of abstractions that define the constrained semantic of the Kokkos parallel programming model.}
\label{fig:abstractions}
\end{figure}

>>>>

Kokkos defines a pure interface for C++ and uses C++ as a programming language. This has multiple reasons. It is our experience that developers of scientific and high-performance computing applications express the demand for C++ and a generic parallel programming model to support their codes. Further, following their estimates, porting efforts towards adapting to vendors, programming models, APIs, and software releases represent a burden worth addressing. Lastly, the possibility to maintain pure C++ codes is appealing from the code development and debugging perspective.

@@

To capture semantic information, that is the~\emph{what}, the~\emph{where}, and the~\emph{how}, Kokkos introduces six abstractions:~\emph{execution spaces},~\emph{execution patterns},~\emph{execution policies},~\emph{memory spaces},~\emph{memory layout}, and~\emph{memory traits}. These abstractions specify semantic information that enables to capture the programmer's intent and allows the runtime to efficiently map the program to any underlying hardware architecture. We list them as follows:
\begin{itemize}
	\item  An execution space is a place where code can be executed. On current hardware architectures, this corresponds to accelerators and CPUs and can include other compute devices in the future. This abstraction supports remote compute devices in distributed memory scenarios as remote execution spaces.
	\item Execution patterns expose the parallel programming paradigm. Supported patterns are the \emph{parallel\_for} loop that executes the loop body in any order a specified amount of times, the \emph{parallel\_reduce} which combines a parallel\_for with a reduction operation, the \emph{parallel\_scan} which combines a parallel\_for operation with a prefix or postfix scan, and \emph{task} which executes a single function potentially in parallel with respect to other tasks or code sections. 
	\item Execution policies shape the iteration space of a loop pattern. A simple execution policy is a range policy. It specifies that the loop body is executed once for each element in a range.
	\item Memory spaces are the places where data resides. They specify the physical locations of data as well as access characteristics. Different physical locations correspond to different device types such as high bandwidth memories, on-die scratch memories, or non-volatile bulk storage. Different logical memory spaces allow for concepts such as memory in the CUDA programming model, which is accessible from the host and the CUDA accelerator. Memory spaces, similarly to execution spaces, conceptually support remote memory locations in distributed-memory scenarios. Furthermore, they encapsulate functionality such as consistency control and persistence scopes.
	\item Layouts express the mapping from array indices to address offsets. By adopting appropriate layouts for memory structures, an application can optimize data access patterns in a given algorithm. If an implementation provides polymorphic layouts (i.e. a data structure can be instantiated at compile or runtime with different layouts), architecture-dependent optimizations can be performed.
	\item Memory traits specify how a data structure is accessed. Traits express usage scenarios such as atomic access, random access, and streaming loads or stores. This allows the programming model to optimize load and store operations.
\end{itemize}

\begin{figure}[t!]
\begin{minted}[fontsize=\small]{c++}
template <class ExecPolicy, class FunctorType>
Kokkos::parallel_for(const std::string& name, 
 const ExecPolicy& policy, 
 const FunctorType& functor);
\end{minted}
\caption{The Kokkos~\emph{parallel\_for} class shows how template metaprogramming allows to specialize types. Its descriptive semantic offers the freedom to optimize execution on modern architectures with multiple degrees of concurrency where an execution policy (\emph{ExecPolicy}) shapes the iteration space accordingly.}
\label{fig:parallelFor}
\end{figure}

@@

\begin{figure}[t!]
\begin{minted}[fontsize=\small]{text}
Constrained Semantics: Patterns (intent),
Spaces, Layouts, Policies and Traits
Parallel Paradigms: Parallel patterns and 
tasking
Expression: Embedded, C++
Control Paradigm: descriptive
\end{minted}
\caption{Semantic capture defined by the Kokkos programming model.}
\label{fig:SemCaptureKokkos}
\end{figure}

@@

\begin{figure}[t!]
\centerline{\includegraphics[width=0.49\textwidth]{img/Abstractions.png}}
\caption{Overview of abstractions that define the constrained semantic of the Kokkos parallel programming model.}
\label{fig:abstractions}
\end{figure}

### 2020-10-13 src/5.Example.tex
Cam-ready version.

In this section we provide an example of an application written in the Kokkos programming model. The code in Figure~\ref{fig:KokkosExample} shows a parallel matrix-vector multiplication implemented as two nested loops. The outer loop, that is the outer~\emph{parallel\_for}, iterates over the rows of the input matrix and of the input vector while the inner loop iterates over elements in each row. Since the inner loop potentially executes in parallel, the update operation of the aggregating variable~\emph{y\_tmp} must be protected against data races. Using the~\emph{parallel\_reduction} pattern, the programming model provides a local variable to that loop iteration. Once the computation finishes, the final result of the reduction operation is stored in the output vector~\emph{y} at the row index~\emph{e}.  
As a first step, a Kokkos application must call ~\emph{Kokkos::initialize}. Initialization and finalization are required and serve the purpose of setting up the underlying runtime. 
Further, the example code shows the use of the ~\emph{Kokkos::View} abstraction.

@@

While this example can only provide a brief insight into Kokkos, we recommend to the interested reader to access further resources. Resources include example applications for each individual Kokkos feature, tutorials with multiple days worth of lectures and exercises, a programming guide, and API documentation. Source code examples as well as the tutorial are located in the Kokkos repository~\cite{KOKKOS_REPO}. The tutorial is intended for students with previous knowledge of one parallel programming model and basic knowledge of C++. It introduces concepts of parallel programming using Kokkos through a series of lectures and hands-on exercises that build upon each other and gradually introduce new concepts. The programming guide is available on-line~\cite{KOKKOS_WIKI}. In chapter~\ref{chap:teaching} we provide further details on best practices obtained from teaching Kokkos as well as list feedback received from lecture attendees.

\begin{figure}
\begin{small}
\begin{minted}[fontsize=\small]{c++}
#include<Kokkos_Core.hpp>
int main(int argc, char* argv[]) {
  Kokkos::initialize(argc,argv);
  {
    Kokkos::View<double*> x("x", M);  
    Kokkos::View<double*> y("y", N);
    Kokkos::View<double**,Kokkos::LayoutRight> 
                                  A("A", N, M);  
    auto x_h = Kokkos::create_mirror_view(x);
    auto y_h = Kokkos::create_mirror_view(y);
    auto A_h = Kokkos::create_mirror_view(A);
    Kokkos::deep_copy(x,x_h);
    Kokkos::deep_copy(A,A_h);

>>>>

In this section, we provide an example of an application written in the Kokkos programming model. The code in Figure~\ref{fig:KokkosExample} shows a parallel matrix-vector multiplication implemented as two nested loops. The outer loop, that is the outer~\emph{parallel\_for}, iterates over the rows of the input matrix and of the input vector while the inner loop iterates over elements in each row. Since the inner loop potentially executes in parallel, the update operation of the aggregating variable~\emph{y\_tmp} must be protected against data races. Using the~\emph{parallel\_reduction} pattern, the programming model provides a local variable to that loop iteration. Once the computation finishes, the final result of the reduction operation is stored in the output vector~\emph{y} at the row index~\emph{e}.  
As a first step, a Kokkos application must call ~\emph{Kokkos::initialize}. Initialization and finalization are required and serve the purpose of setting up the underlying runtime. 
Further, the example code shows the use of the ~\emph{Kokkos::View} abstraction.

@@

While this example can only provide a brief insight into Kokkos, we recommend to the interested reader to access further resources. Resources include example applications for each Kokkos feature, tutorials with multiple days worth of lectures and exercises, a programming guide, and API documentation. Source code examples as well as the tutorial are located in the Kokkos repository~\cite{KOKKOS_REPO}. The tutorial is intended for students with previous knowledge of one parallel programming model and basic knowledge of C++. It introduces concepts of parallel programming using Kokkos through a series of lectures and hands-on exercises that build upon each other and gradually introduce new concepts. The programming guide is available on-line~\cite{KOKKOS_WIKI}. In chapter~\ref{chap:teaching} we provide further details on best practices obtained from teaching Kokkos as well as list feedback received from lecture attendees.

\begin{figure}[t!]
\begin{small}
\begin{minted}[fontsize=\small]{c++}
#include<Kokkos_Core.hpp>
int main(int argc, char* argv[]) {
  Kokkos::initialize(argc,argv);
  {
    Kokkos::View<double*> x("x", M);  
    Kokkos::View<double*> y("y", N);
    Kokkos::View<double**,Kokkos::LayoutRight> 
                                  A("A", N, M);  
    auto x_h = Kokkos::create_mirror_view(x);
    auto y_h = Kokkos::create_mirror_view(y);
    auto A_h = Kokkos::create_mirror_view(A);
    Kokkos::deep_copy(x,x_h);
    Kokkos::deep_copy(A,A_h);

### 2020-10-13 src/8.Related.tex
Cam-ready version.

C++ increasingly supports parallel constructs. It is becoming increasingly possible to teach parallelism in C++ and relying on modern C++ compilers. However, at the time of publication, C++ does not support accelerators or many of the complexities of parallel computing related to accelerator programming. In courses that do not target accelerators, this can be a relevant option. Over the past decade the Kokkos programming model has contributed to a number of current and near-future ISO-C++ features, including \emph{atomic\_ref}\cite{wg21_p0019}, \emph{mdspan}\cite{MDSPAN}, and executors.  In particular, the design of C++ executors, which are similar to Kokkos execution spaces, were influenced by this programming model. As the demand to extract performance from increasingly deep and increasingly asynchronous software stacks across a wide variety of domains increases, many C++ experts expect executors to become a central abstraction in any performance-oriented software stack. 
RAJA~\cite{Raja} is a C++ abstraction library with similarities to Kokkos. The important difference however is that Kokkos is descriptive, while RAJA is prescriptive. In Kokkos, the programming model determines how an application is mapped to the underlying hardware. RAJA provides functionality that exposes hardware details and relies on the developer to try different strategies to map kernels to architectures. RAJA is a viable option for class education in which the educator would like to expose students to the semantics of different programming models without exposing them to the detailed syntax of those models.
DPC++\cite{DPCPP} is a model developed by Intel\textsuperscript{\textregistered} for expressing parallelism across Intel hardware architectures. In addition to sharing a descriptive philosophy with Kokkos, it also shares parallel patterns. Unlike Kokkos however, it targets Intel architectures only and is not vendor-agnostic. A course targeting Intel architectures, DPC++ could be a valid option.

>>>>

C++ increasingly supports parallel constructs. It is becoming increasingly possible to teach parallelism in C++ and relying on modern C++ compilers. However, at the time of publication, C++ does not support accelerators or many of the complexities of parallel computing related to accelerator programming. In courses that do not target accelerators, this can be a relevant option. Over the past decade, the Kokkos programming model has contributed to a number of current and near-future ISO-C++ features, including \emph{atomic\_ref}\cite{wg21_p0019}, \emph{mdspan}\cite{MDSPAN}, and executors.  In particular, the design of C++ executors, which are similar to Kokkos execution spaces, were influenced by this programming model. As the demand to extract performance from increasingly deep and increasingly asynchronous software stacks across a wide variety of domains increases, many C++ experts expect executors to become a central abstraction in any performance-oriented software stack. 
RAJA~\cite{Raja} is a C++ abstraction library with similarities to Kokkos. The important difference however is that Kokkos is descriptive, while RAJA is prescriptive. In Kokkos, the programming model determines how an application is mapped to the underlying hardware. RAJA provides functionality that exposes hardware details and relies on the developer to try different strategies to map kernels to architectures. RAJA is a viable option for class education in which the educator would like to expose students to the semantics of different programming models without exposing them to the detailed syntax of those models.
DPC++\cite{DPCPP} is a model developed by Intel\textsuperscript{\textregistered} for expressing parallelism across Intel hardware architectures. In addition to sharing a descriptive philosophy with Kokkos, it also shares parallel patterns. Unlike Kokkos however, it targets Intel architectures only and is not vendor-agnostic. A course targeting Intel architectures, DPC++ could be a valid option.
